Modelling writing hesitations in text writing as finite mixture process

### 2023-05-05 report/results.Rmd
Add tutorial and draft results

```{r include = F}
library(tidyverse)
library(kableExtra)
library(janitor)
library(brms)
library(polspline)
library(patchwork)
library(rmdformats)
library(gridExtra)
library(grid)
library(gtable)
library(gridExtra)
library(scales)
#library(caret)
#library(InformationValue)

# Calculate Bayes Factor
BF <- function(ps, prior_sd = 1){
  fit_posterior <- logspline(ps) 
  posterior <- dlogspline(0, fit_posterior) # Height of the posterior at 0 
  prior <- dnorm(0, 0, prior_sd) # Height of the prior at 0
  BF10 <- prior / posterior 
  return(BF10)
}

logit <- function(p) log(p / (1-p))
ilogit <- function(x) 1 / (1 + exp(-x))
inv_logit <- function(logit) exp(logit) / (1 + exp(logit))

se_bin <- function(x) sqrt((mean(x, na.rm = T)*(1 - mean(x, na.rm = T)))/length(x)) # se for binary data

lower <- function(x) quantile(x, prob = .025)
upper <- function(x) quantile(x, prob = .975)

# Remove leading zeros and round numbers
dezero <- function(x, dp){ # dp is decimal places
  fmt = paste0('%.',dp,'f')
  x = sprintf(fmt,x)
  x = str_replace(x, '(-|^)0\\.','\\1\\.')
  return(x)
}
dezero_plot <- function(x) dezero(x, 1)

@@

@@

Data were analysed in Bayesian mixed effects models [@gelman2014;@mcelreath2016statistical]. The R [@R-base] package rstan [@rstan] was used to interface with the probabilistic programming language Stan [@carpenter2016stan] which was used to implement all models. Models were fitted with weakly informative priors [see @mcelreath2016statistical], and run with 10,000 iterations on 3 chains with a warm-up of 5,000 iterations and no thinning. Model convergence was confirmed by the Rubin-Gelman statistic ($\hat{R}$ = 1) [@gelman1992] and inspection of the Markov chain Monte Carlo chains.

@@

# Data processing

TODO: Add a description with general data filtering and a table with how many data were removed from the analysis by study.

# C2L1

The C2L1 data set comprises data Norwegian 6th graders--*N*=133, mean age 11 years 10 months--published in @ronneberg2022process. The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography.

## Model comparisons

```{r predictionsc2l1, fig.cap="C2L1 data. Comparison of 100 simulated (predicted) sets of data to observed data illustated by model. For illustration the x-axis was truncated at 2,000 msecs."}

files <- list.files("../stanout/c2l1", pattern = "sims", full.names = T)
sims <- read_csv(files) %>% 
  pivot_longer(cols = c(y_obs, y_tilde)) %>% 
  mutate(across(name, recode_factor, 
                y_tilde = "Simulated data",
                y_obs = "Observed data",
                .ordered = T)) %>% 
  filter(name ==  "Simulated data" |
         (name == "Observed data" & sim_idx == 1))

max_iki <- 2000

plot_mog <- sims %>%
  filter(model == "mog", value < max_iki) %>% 
  mutate(across(model, recode, mog = "Bimodal log-normal")) %>% 
  ggplot(aes(x = value, 
             group = interaction(sim_idx,name), 
             colour = name)) +
  geom_density() +
  facet_wrap(~model) +
  scale_colour_manual(values = c("firebrick3", "black")) +
  scale_x_continuous(labels = scales::comma) +
  labs(colour = "") +
  theme(axis.title = element_blank())

plot_lmm <- sims %>%
  filter(model == "lmm", value < max_iki) %>% 
  mutate(across(model, recode, lmm = "Unimodal log-normal")) %>% 
  ggplot(aes(x = value, 
             group = interaction(sim_idx,name), 
             colour = name)) +
  geom_density() +
  facet_wrap(~model) +
  scale_colour_manual(values = c("firebrick3", "black")) +
  scale_x_continuous(labels = scales::comma) +
  labs(colour = "") +
  theme(axis.title = element_blank())

plot_lmmgaus <- sims %>%
  filter(model == "lmmgaus", value < max_iki) %>% 
  mutate(across(model, recode, lmmgaus = "Unimodal normal")) %>% 
  ggplot(aes(x = value, 
             group = interaction(sim_idx,name), 
             colour = name)) +
  geom_density() +
  facet_wrap(~model) +
  scale_colour_manual(values = c("firebrick3", "black")) +
  scale_x_continuous(labels = scales::comma) +
  labs(colour = "") +
  theme(axis.title = element_blank())

plot_lmmuneqvar <- sims %>%
  filter(model == "lmmuneqvar", value < max_iki) %>% 
  mutate(across(model, recode, lmmuneqvar = "Unimodal log-normal (unequal variance)")) %>% 
  ggplot(aes(x = value, 
             group = interaction(sim_idx,name), 
             colour = name)) +
  geom_density() +
  facet_wrap(~model) +
  scale_colour_manual(values = c("firebrick3", "black")) +
  scale_x_continuous(labels = scales::comma) +
  labs(colour = "") +
  theme(axis.title = element_blank())

plots <- plot_lmmgaus + plot_lmm + plot_lmmuneqvar + plot_mog +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = 'A')

### Out-of-samples cross-validation

```{r}
files <- list.files("../stanout/c2l1", pattern = "^modelcomp", full.names = T)
table <- purrr::map_dfr(files, read_csv) %>% 
  mutate(across(location, replace_na, "overall")) %>% 
  select(location, model, elpd_diff, se_diff, elpd_loo, se_elpd_loo) %>%
  mutate(across(where(is.numeric), round, 0),
         across(where(is.numeric), format, big.mark = ","),
         across(everything(), str_trim)) %>%
  transmute(across(location, ~str_c("Text location: ",.)),
            across(model, recode, lmm = "Unimodal log-normal",
                                  lmmgaus = "Unimodal normal",
                                  mog = "Bimodal log-normal",
                                  lmmuneqvar = "Unimodal log-normal (unequal variance)"),
            across(elpd_diff, str_c, " (", se_diff, ")"),
            across(elpd_loo, str_c, " (", se_elpd_loo, ")")) %>% 
  mutate(across(elpd_diff, str_replace, "0 \\(0\\)", "--")) %>% 
  pivot_wider(names_from = location, values_from = contains("elpd")) %>% 
  select(Model = model, 
         ends_with("overall"), 
         ends_with("sent"), 
         ends_with("beforeword"), 
         ends_with("withinword"))

@@

```{r c2l1, results='asis'}
names(table)[-1] <- rep(c("$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$"),4)

kable(table, caption = "C2L1 data. Model comparisons. The top row shows the models with the highest predictive performance. Standard error is shown in parentheses.",
        align =c("l", rep("r", 8))) %>%
  kable_styling(position = "center", 
                full_width = T, 
                font_size = 11) %>%
  footnote("$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.") %>%
  add_header_above(c(" " = 1, 
                     "All transitions" = 2,
                     "Before sentence" = 2, 
                     "Before word" = 2,
                     "Within word" = 2)) %>%
  column_spec(1, width = "8em")

@@

## Posterior parameter estimates of mixture model

```{r c2l1post, fig.cap="C2l1. Posterior parameter distribution"}
file <- list.files(path = "../stanout/c2l1", pattern = "mog.csv$",full.names = T)
ps <- read_csv(file) %>% 
  pivot_wider(names_from = param, values_from = value) %>% 
  unnest(cols = c(beta, delta, prob)) %>% 
  mutate(across(delta, ~.+beta),
         across(c(delta, beta), exp),
         across(delta, ~.-beta),
         across(location, str_replace, " ", "\n")) %>% 
  pivot_longer(beta:prob) %>% 
  group_by(location, name) %>% 
  summarise(across(value, list(mean = mean, 
                               lower = lower, 
                               upper = upper), .names = "{.fn}"))

beta <- filter(ps, name == "beta") %>% 
  mutate(name = beta_label) %>% 
  ggplot(aes(y = mean, 
             ymin = lower, 
             ymax = upper,
             x = location)) +
  geom_point(size = 3, position = position_dodge(.25)) +
  geom_errorbar(width = 0, position = position_dodge(.25)) +
  facet_grid(~name) +
  scale_shape_manual(values = 1:4) +
  scale_colour_viridis_d(end = .7) +
  scale_y_continuous(labels = scales::comma) +
  labs(y = "Posterior estimate with 95% PIs",
       x = "",
       colour = "",
       shape = "") +
  guides(colour=guide_legend(nrow=2,byrow=TRUE),
         shape = guide_legend(nrow=2,byrow=TRUE))

delta <- filter(ps, name == "delta") %>% 
  mutate(name = delta_label) %>% 
  ggplot(aes(y = mean, 
             ymin = lower, 
             ymax = upper,
             x = location)) +
  geom_point(size = 3, position = position_dodge(.25)) +
  geom_errorbar(width = 0, position = position_dodge(.25)) +
  facet_grid(~name) +
  scale_shape_manual(values = 1:4) +
  scale_colour_viridis_d(end = .7) +
  scale_y_log10(labels = scales::comma) +
  labs(y = "",
       x = "Transition location",
       colour = "",
       shape = "") +
  theme(axis.title.y = element_blank()) +
  guides(colour=guide_legend(nrow=2,byrow=TRUE),
         shape = guide_legend(nrow=2,byrow=TRUE))

theta <- filter(ps, name == "prob") %>% 
  mutate(name = theta_label) %>% 
  ggplot(aes(y = mean, 
             ymin = lower, 
             ymax = upper,
             x = location)) +
  geom_point(size = 3, position = position_dodge(.25)) +
  geom_errorbar(width = 0, position = position_dodge(.25)) +
  facet_grid(~name) +
  scale_shape_manual(values = 1:4) +
  scale_colour_viridis_d(end = .7) +
  scale_y_continuous(labels = dezero_plot, limits = c(0, 1)) +
  labs(y = "",
       x = "",
       colour = "",
       shape = "") +
  theme(axis.title.y = element_blank()) +
  guides(colour=guide_legend(nrow=2,byrow=TRUE),
         shape = guide_legend(nrow=2,byrow=TRUE))

@@

# CATO

## Model comparisons

files <- list.files("../stanout/cato", pattern = "sims", full.names = T)
sims <- read_csv(files) %>% 
  pivot_longer(cols = c(y_obs, y_tilde)) %>% 
  mutate(across(name, recode_factor, 
                y_tilde = "Simulated data",
                y_obs = "Observed data",
                .ordered = T)) %>% 
  filter(name ==  "Simulated data" |
         (name == "Observed data" & sim_idx == 1))

max_iki <- 2000

plot_mog <- sims %>%
  filter(model == "mog", value < max_iki) %>% 
  mutate(across(model, recode, mog = "Bimodal log-normal")) %>% 
  ggplot(aes(x = value, 
             group = interaction(sim_idx,name), 
             colour = name)) +
  geom_density() +
  facet_wrap(~model) +
  scale_colour_manual(values = c("firebrick3", "black")) +
  scale_x_continuous(labels = scales::comma) +
  labs(colour = "") +
  theme(axis.title = element_blank())

plot_lmm <- sims %>%
  filter(model == "lmm", value < max_iki) %>% 
  mutate(across(model, recode, lmm = "Unimodal log-normal")) %>% 
  ggplot(aes(x = value, 
             group = interaction(sim_idx,name), 
             colour = name)) +
  geom_density() +
  facet_wrap(~model) +
  scale_colour_manual(values = c("firebrick3", "black")) +
  scale_x_continuous(labels = scales::comma) +
  labs(colour = "") +
  theme(axis.title = element_blank())

plot_lmmgaus <- sims %>%
  filter(model == "lmmgaus", value < max_iki) %>% 
  mutate(across(model, recode, lmmgaus = "Unimodal normal")) %>% 
  ggplot(aes(x = value, 
             group = interaction(sim_idx,name), 
             colour = name)) +
  geom_density() +
  facet_wrap(~model) +
  scale_colour_manual(values = c("firebrick3", "black")) +
  scale_x_continuous(labels = scales::comma) +
  labs(colour = "") +
  theme(axis.title = element_blank())

plot_lmmuneqvar <- sims %>%
  filter(model == "lmmuneqvar", value < max_iki) %>% 
  mutate(across(model, recode, lmmuneqvar = "Unimodal log-normal (unequal variance)")) %>% 
  ggplot(aes(x = value, 
             group = interaction(sim_idx,name), 
             colour = name)) +
  geom_density() +
  facet_wrap(~model) +
  scale_colour_manual(values = c("firebrick3", "black")) +
  scale_x_continuous(labels = scales::comma) +
  labs(colour = "") +
  theme(axis.title = element_blank())

plots <- plot_lmmgaus + plot_lmm + plot_lmmuneqvar + plot_mog +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = 'A')

### Out-of-samples cross-validation

```{r}
files <- list.files("../stanout/cato", pattern = "^modelcomp", full.names = T)
table <- purrr::map_dfr(files, read_csv) %>% 
  mutate(across(location, replace_na, "overall")) %>% 
  select(location, model, elpd_diff, se_diff, elpd_loo, se_elpd_loo) %>%
  mutate(across(where(is.numeric), round, 0),
         across(where(is.numeric), format, big.mark = ","),
         across(everything(), str_trim)) %>%
  transmute(across(location, ~str_c("Text location: ",.)),
            across(model, recode, lmm = "Unimodal log-normal",
                                  lmmgaus = "Unimodal normal",
                                  mog = "Bimodal log-normal",
                                  lmmuneqvar = "Unimodal log-normal (unequal variance)"),
            across(elpd_diff, str_c, " (", se_diff, ")"),
            across(elpd_loo, str_c, " (", se_elpd_loo, ")")) %>% 
  mutate(across(elpd_diff, str_replace, "0 \\(0\\)", "--")) %>% 
  pivot_wider(names_from = location, values_from = contains("elpd")) %>% 
  select(Model = model, 
         ends_with("overall"), 
         ends_with("sent"), 
         ends_with("beforeword"), 
         ends_with("withinword"))

@@

>>>>

```{r include = F}
library(tidyverse)
library(kableExtra)
library(janitor)
library(brms)
library(polspline)
library(patchwork)
library(rmdformats)
library(gridExtra)
library(grid)
library(gtable)
library(gridExtra)
library(scales)
library(ggthemes)
source("../scripts/functions.R")
source("scripts/get_pdens_plot.R")

@@

# Introduction

```{r eval = F}
Take intro from Mark's draft of SPL2 paper

Different views on pausing in text production, largely representing typing as a unimodal or bimodal process

- pauses are represented in the right tail of the distribution (unimodal) explaining the long positive skew and therefore possibly making a log transformation redundant (there are people arguing against log transformation of keystrokes but I don't understand their argument)
- as pauses are more likely at linguistic edges, the variance of the distribution is going to be larger
- pauses are represented as an independent distribution / process
  - short distribution: fluent key transitions only limited by the ability to move fingers (same across text locations)
  - short distribution: processes involved in cascade are different when lexical and syntactic planning is involved (differences across text locations)

# Model definitions

The models presented in the following can be divided into two general groups. The first three models are largely akin to models typically used in the literature. By this we mean models that assume a uni-modal process that generates keystroke data as is incorporated in statistical models such as analysis of variance and linear mixed-effects models. Second, the last two models model keystroke intervals as a combination of two weighted processes of which one presents a smooth information flow from mind into finger; the other component is more important as it represents moments at which the information flow was interrupted leading to longer latencies. The latter two models map directly on the idea of a cascading model of writing.

## Uni-modal Gaussian

We start with a Gaussian mixed-effects models similar to a standard analysis of variance. We describe the process that generates each iki $i$ as normal Gaussian distribution $\text{N}()$ characterised by a mean $\mu$ and a standard deviation $\sigma_\text{e}^2$. The mean can be decomposed into $\beta$ and $\text{participant}_i$. $\beta$ will be allowed to take on a different value for each transition location. Average participant-ikis are allowed to deviate from the average which is achieved by assuming a normal distribution for participant deviations distributed around 0 with a standard deviation $\sigma_\text{p}^2$.

$$
\begin{align}
\text{iki}_i & \sim \text{N}(\mu_i, \sigma_\text{e}^2)\\
\text{where: } & \mu_i = \beta_\text{location[i]} + \text{participant}_i\\
& \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) 
\end{align}
$$

## Uni-modal log-Gaussian

This model is largely identical to the previous model, except we use a log-normal distribution instead of a normal distribution. The advantage of using a log-normal distribution is two-fold: (1) the log-normal distribution has a lower bound of zero. For our data we generally consider negative ikis as mistakes which occasionally occur when the next key was pressed before the current key. Other than that keystroke intervals are constrained by a persons ability to move their fingers and keyboard polling. (2) the log-scale is known to be a better match for data from human behaviour, in particular motor responses. In particular, a normal distribution assumes that units are linearly scaled. For example, a 50 msecs difference is the same between 100 msecs and 150 msecs than it is between 5 secs and 5 secs 50 msecs (i.e. 5,050 msecs). This is not necessarily plausible for keystroke data. We would assume that differences that are due to motor activity (typing a high frequency bigram vs a low frequency bigram) are smaller than difference that are due to high levels of cognitive activity (retrieving a word in your L1 or L2). Log-normal distributions are a natural way of scaling units so that a 50 msecs difference on the lower end of the iki scale (motor activity) is more meaningful than a 50 msecs difference on the upper end of the iki scale (retrieving words, planning sentences).

The model can be described like this:

$$
\begin{align}
\text{iki}_i & \sim \text{logN}(\mu_i, \sigma_\text{e}^2) \\
\text{where: } &
\mu_i = \beta_\text{location[i]} + \text{participant}_i\\
& \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) 
\end{align}
$$

## Uni-modal unequal-variance log-Gaussian

$$
\begin{align}
\text{iki}_i & \sim \text{logN}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
\text{where: } & \mu_i = \beta_\text{location[i]} + \text{participant}_i\\
 & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) 
\end{align}
$$

## Bi-modal log-Gaussian (constrained)

This models extends the intuition from the previous model that higher levels of activation lead to longer pauses. Instead of assuming that there is one process that underlies the generation of ikis, we assume there are two. (1) activation can flow into keystrokes without interrupts. These fluent keystroke transitions are merely limited by a person's ability to move their finger and will be captured but the parameter $\beta$. In principle, there are no differences for fluent key-transitions between transition location. The next model with loosen this assumption (hence, "unconstrained model"). (2) difficulty in the activation flow leads to pauses when fingers have to catch-up with cognitive activity, when spelling, words, or contents couldn't be retrieved in time. The size of these pauses will depend on the reason for delays which is typically associated with transition locations (contents are typically planned before sentences, words are retrieved before they are typed and spelling difficulty typically occurs when typing a word). Pauses will be captured by two model parameters: (1) the slowdown for these hesitant transitions will be captured by $\delta$ which is the deviation compared to normal typing intervals (constrained to be positive). (2) the frequency of hesitant transitions will be captured by $\theta$ for each level of a categorical predictor.

$$
\begin{align}
\text{iki}_{i} & \sim \theta_\text{location[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + \text{participant}_i, \sigma_{e'_\text{location[i]}}^2) + \\
  & (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta + \text{participant}_i, \sigma_{e}^2)\\
\text{where: }
		&\delta \sim \text{N}(0,1)\\
\text{constraint: } & \delta > 0\\
		& \sigma_{e'}^2 > \sigma_{e}^2
\end{align}
$$
This models takes into account two sources of participant-specific error: (1) each participant has an individual fluent typing as in the previous models; (2) each participant has in individual hesitation frequency that differs across levels of the categorical predictor.

## Bi-modal log-Gaussian (unconstrained)

This model is identical to the previous model with one exception. The distribution of fluent keystroke transitions captured by $\beta$ was fixed to be the same across transition locations is the previous model. In other words, the mean $\beta$ and it's standard deviation $\sigma_{e}^2$ was the same for before-sentence, before-word, and within-word transitions. This means, because of the naturally larger number of within-word transitions, the posterior is dominated by within-word transitions.

In this model we will loosen this constraint and allow $\beta$ and $\sigma_{e}^2$ to vary by transition-location.

$$
\begin{align}
\text{iki}_{i} &\sim \theta_\text{location[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + \text{participant}_i, \sigma_{e'_\text{location[i]}}^2) + \\
  & (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + \text{participant}_i, \sigma_{e_\text{location[i]}}^2)\\
	\text{where: } & \delta \sim \text{N}(0,1)\\
\text{constraint: } & \delta > 0\\
		& \sigma_{e'}^2 > \sigma_{e}^2
\end{align}
$$

@@

Data were analysed in Bayesian mixed effects models [@gelman2014;@mcelreath2016statistical]. The R [@R-base] package rstan [@rstan] was used to interface with the probabilistic programming language Stan [@carpenter2016stan] which was used to implement all models. Models were fitted with weakly informative priors [see @mcelreath2016statistical], and run with 20,000 iterations on 3 chains with a warm-up of 10,000 iterations and no thinning. Model convergence was confirmed by the Rubin-Gelman statistic ($\hat{R}$ = 1) [@gelman1992] and inspection of the Markov chain Monte Carlo chains.

# Datasets in overview

Five dataset with keystroke data from text production were used for analysis. An overview can be found in Table \@ref(tab:datasets).

```{r datasets, results='asis'}
tibble(Dataset = c("C2L1", "CATO", "SPL2", "PLanTra", "LIFT"),
                Source = c("@ronneberg2022process",
                              "@torrance2016adolescent",
                              "@torrance",
                              "@rossetti2022s",
                              "@nina_vandermeulen_2020_3893538"),
                Keylogger = c("EyeWrite", 
                           "EyeWrite", 
                           "CyWrite", 
                           "InputLog", 
                           "InputLog"),
                `Writing task` = c(
                                 "Argumentative essays", 
                                 "Expository texts",
                                 "Argumentative essays",
                                 "Text simplification",
                                 "Synthesis"),
                n_ppt = c(126, 26*2, 39, 47, 658),
#                n_texts = c(1, 2, 2, 2, NA),
#                n_sentences = c(),
#                n_words = c(),
                conditions = c(" ", 
                               "weak decoders / control; masked / unmasked",
                               "write in L1 / L2",
                               "pre / post test trained in plain language principles and control", 
                               "Various topics and genres"),
                `Mean age` = c(11.8, 16.9, 20.6, 23, 16.95),
                 Language = c( "Norwegian", 
                               "Norwegian", 
                               "English (L1) / Spanish (L2)", 
                               "English (L2)", 
                               "Dutch")) %>% 
  kable(caption = "Datasets in brief.") %>%
  kable_styling(position = "center", 
                full_width = T, 
                font_size = 12)

```{r eval = F}
Data reduction using random samples of subsets of participants for larger data sets

Also for larger data sets, only used participants that provided a min number of sentences

Locations with example from Marks

Removed transitions that were followed by an editing operation

Extremely short and long intervals

Add a description with general data filtering and a table with how many data were removed from the analysis by study.

# Out-of-samples cross-validation

@@

Results for all data sets are shown in Table \@ref(tab:loos). For all data sets we found the same pattern. The mixture of log-normal distributions provided a substantially better fit than uni-modal distribution models. The unconstrained version of the mixture of log-normal distributions rendered a higher predictive performance than the constrained version that does not allow the distribution of short keystroke-intervals to vary across conditions.

```{r}
files <- list.files(
  str_c("../stanout/", 
  c("lift", "spl2", "spl2_shift", "plantra", "cato", "c2l1")), 
  pattern = "model.+.csv", 
  full.names = T)

table <- purrr::map_dfr(files, ~read_csv(.) %>% 
                          mutate(dataset = .x)) %>% 
  select(dataset, model, elpd_diff, se_diff, elpd_loo, se_elpd_loo) %>%
  mutate(across(dataset, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
         across(where(is.numeric), round, 0),
         across(where(is.numeric), format, big.mark = ","),
         across(everything(), str_trim)) %>%
  transmute(
     across(dataset, recode, 
                "cato" = "CATO",
                "spl2" = "SPL2", 
                "spl2_shift" = "SPL2 (shift + C)", 
                "plantra" = "PLanTra",
                "lift" = "LIFT",
                "c2l1" = "C2L1"),
    across(model, recode, lmm = "Unimodal log-normal",
                          lmmgaus = "Unimodal normal",
                          mogbetaconstr = "Bimodal log-normal (constrained)",
                          mogbetacontr = "Bimodal log-normal (constrained)",
                          mogbetaunconstr = "Bimodal log-normal (unconstrained)",
                          lmmuneqvar = "Unimodal log-normal (unequal variance)"),
            across(elpd_diff, str_c, " (", se_diff, ")"),
            across(elpd_loo, str_c, " (", se_elpd_loo, ")")) %>% 
  mutate(across(elpd_diff, str_replace, "0 \\(0\\)", "--")) %>% 
  pivot_wider(names_from = dataset, values_from = contains("elpd")) %>% 
  select(Model = model, ends_with("CATO"), ends_with("C2L1"), ends_with("LIFT"), ends_with("PLanTra"), ends_with("SPL2"), ends_with("+ C)"))

@@

```{r loos, results='asis'}
names(table)[-1] <- rep(c("$\\Delta\\widehat{elpd}$", "$\\widehat{elpd}$"), 6)

kable(table, 
      caption = "Model comparisons. The top row shows the models with the highest predictive performance. Standard error is shown in parentheses.",
        align =c("l", rep("r", 12))) %>%
  kable_styling(position = "center", 
                full_width = T, 
                font_size = 11) %>%
  footnote("$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.") %>%
  add_header_above(c(" " = 1, 
                     "CATO" = 2,
                     "CL21" = 2, 
                     "LIFT" = 2,
                     "PLanTra" = 2,
                     "SPL2" = 2,
                     "SPL2 (shift + C)" = 2)) %>%
  column_spec(1, width = "8em")

@@

# Cross-data set comparisons

```{r}
files <- list.files(str_c("../stanout/", 
                          c("lift", "spl2", "plantra", "cato", "c2l1")), 
                    pattern = "mogbetaun.+.csv", 
                    full.names = T)

ps <- purrr::map_dfr(files, ~read_csv(.x) %>% 
  mutate(data = .x)) %>% 
  mutate(across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
         across(data, recode_factor, 
                "cato" = "CATO (non-dyslexic\nunmasked)",
                "spl2" = "SPL2 (L1)", 
#                "spl2 (shift + C)" = "SPL2 (L1; shift + C)", 
                "plantra" = "PLanTra",
                "lift" = "LIFT",
                "c2l1" = "C2L1",
                .ordered = TRUE)) %>% 
  filter(!(str_detect(data, "SPL2") & lang == "ES"),
         !(str_detect(data, "CATO") & group == "dyslexic"),
         !(str_detect(data, "CATO") & task == "masked")) %>% 
  pivot_wider(names_from = param, 
              values_from = value) %>% 
  unnest(cols = beta:theta) %>% 
  mutate(across(c(beta, beta2), exp),
         delta = beta2 - beta,
         across(location, str_replace, " ", "\n")) %>% 
  pivot_longer(beta:theta) %>% 
  summarise(across(value, 
                   list(mean = mean, 
                        lower = lower, 
                        upper = upper),
                   .names = "{.fn}"), .by = c(data, location, name)) %>% 
  filter(name %in% c("beta", "delta", "prob"))
```

## Cross-data set visualisation

The model estimates for the mixture-model with the highest predictive performance are shown in Figure \@ref(fig:crossstudypost). In this visualisation we ignore dataset specific conditions that are presented in detail below.

```{r crossstudypost, fig.cap="Across studies. Posterior parameter distribution"}
posd <- position_dodge(.65)
dotsize <- 2.5
grouplabel <- "Data set"
beta <- filter(ps, name == "beta") %>% 
  mutate(name = beta_label) %>% 
  ggplot(aes(y = mean, 
             ymin = lower, 
             ymax = upper,
             x = location,
             shape = data,
             group = interaction(data))) +
  geom_line(size = .5, 
            alpha = .35, 
            position = posd) +
  geom_errorbar(width = .2, alpha = .75, position = posd) +
  geom_point(aes(colour = data), 
            size = dotsize, 
            position = posd) +
  facet_grid(~name) +
  scale_shape_manual(values = c(1, 4, 7, 8, 9, 11)) +
  scale_colour_colorblind() +
  scale_y_log10(labels = scales::comma) +
  labs(y = "Posterior estimate with 95% PIs",
       x = "",
       colour = grouplabel,
       shape = grouplabel) +
  guides(colour=guide_legend(nrow=2,byrow=TRUE),
         shape = guide_legend(nrow=2,byrow=TRUE))

delta <- filter(ps, name == "delta") %>% 
  mutate(name = delta_label) %>% 
  ggplot(aes(y = mean, 
             ymin = lower, 
             ymax = upper,
             x = location,
             shape = data,
             group = interaction(data))) +
  geom_line(size = .5, alpha = .35, position = posd) +
  geom_errorbar(width = .2, alpha = .75, position = posd) +
  geom_point(aes(colour = data), 
            size = dotsize, 
            position = posd) +
  facet_grid(~name) +
  scale_shape_manual(values = c(1, 4, 7, 8, 9, 11)) +
  scale_colour_colorblind() +
  scale_y_log10(labels = scales::comma) +
  labs(y = "",
       x = "Transition location",
       colour = grouplabel,
       shape = grouplabel) +
  theme(axis.title.y = element_blank()) +
  guides(colour=guide_legend(nrow=2,byrow=TRUE),
         shape = guide_legend(nrow=2,byrow=TRUE))

theta <- filter(ps, name == "prob") %>% 
  mutate(name = theta_label) %>% 
  ggplot(aes(y = mean, 
             ymin = lower, 
             ymax = upper,
             x = location,
             shape = data,
             group = interaction(data))) +
  geom_line(size = .5, alpha = .35, position = posd) +
  geom_errorbar(width = .2, alpha = .75, position = posd) +
  geom_point(aes(colour = data), 
            size = dotsize, 
            position = posd) +
  facet_grid(~name) +
  scale_shape_manual(values = c(1, 4, 7, 8, 9, 11)) +
  scale_colour_colorblind() +
  scale_y_continuous(labels = dezero_plot, limits = c(0, 1)) +
  labs(y = "",
       x = "",
       colour = grouplabel,
       shape = grouplabel) +
  theme(axis.title.y = element_blank()) +
  guides(colour=guide_legend(nrow=2,byrow=TRUE),
         shape = guide_legend(nrow=2,byrow=TRUE))

@@

## Effect of transition location

It's generally believed that pausing is associated with syntactic edges such that more and longer pauses are predicted for key transitions at larger syntactic edges, i.e. before sentence > before word > within word. We have evaluated the differences between transition locations for all data sets. The results are shown in Table \@ref(tab:loceffect).

Results are largely consistent across data sets (with caveats) but differ, to some extent, from what the literature would predict. In line with the literature hesitations are more frequent before words than within words. Also hesitations are longer at before-sentence transitions compared to before-word transitions (except dataset C2L1) compared to within-word transitions (except dataset LIFT). However, our results do not support that writers pause more frequently at before-sentence locations compared to before-word locations (except for dataset SPL2; this also shows that more pauses at before-sentence locations can not be explained on the basis of multi-key combinations for sentence-initial capitalisation). Also, we observe that even fluent key-transitions are slower at before-word locations compared to within-word locations but there is generally not difference for fluent transitions for before-sentence transitions compared to before-word transitions (except for dataset SPL2).

The datasets differ to the extent that sentence-initial key transitions do (PLanTra, LIFT) or do not (CATO, C2L1, SPL2) include the character following the shift key for capitalisation. In other words, the pause before sentences may sum across two key intervals, namely `_^[shift]^C` but only involves one keyintervals, namely `_^[shift]`. For the SPL2 dataset, we calculated location effects for sentence-initial transitions that do and do not involve the shift-to-key transition. The results were the same for the transition location effects. A comparison that is untangling the effects of the multi-keycombination on the mixture-model estimates can be found in Table \@ref(tab:shiftcellmeans). In short, the duration of fluent transitions and the hesitation slowdown are affected but not the hesitation probability.

In conclusion, while pauses tend to be longer before sentences they are not more frequent than before words.

```{r loceffect, results='asis'}
files <- list.files(str_c("../stanout/", 
                          c("lift", "spl2", "spl2_shift", "plantra", "cato", "c2l1")), 
                    pattern = "mogbetaun.+.csv", 
                    full.names = T)

ps_loc_diffs <- purrr::map_dfr(files, ~read_csv(.x) %>% 
  mutate(data = .x)) %>% 
  mutate(across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
         across(data, recode_factor, 
                "cato" = "CATO (non-dyslexic\nunmasked)",
                "spl2" = "SPL2 (L1)", 
                "spl2_shift" = "SPL2 (L1; shift + C)", 
                "plantra" = "PLanTra",
                "lift" = "LIFT",
                "c2l1" = "C2L1",
                .ordered = TRUE)) %>% 
  filter(!(str_detect(data, "SPL2") & lang == "ES"),
         !(str_detect(data, "CATO") & group == "dyslexic"),
         !(str_detect(data, "CATO") & task == "masked"),
         param %in% c("beta", "delta", "theta")) %>%
  select(param, location, data, value) %>% 
  mutate(across(value, ~ifelse(param == "theta", .*-1, .))) %>% 
  pivot_wider(names_from = location, values_from = value) %>%
  unnest(-c(data, param)) %>% 
  mutate(diff.1 = `before sentence` - `before word`,
         diff.2 = `before word` - `within word`) %>% 
  summarise(across(starts_with("diff"), 
                   list(mean = mean, lower = lower, upper = upper, BF = BF)),
            .by = c(data, param))

tmp <- ps_loc_diffs %>% 
  pivot_longer(starts_with("diff"), names_to = c("diffid", ".value"), names_sep = "_") %>% 
  mutate(across(mean, ~pmap_chr(list(., lower, upper, 2), PI)),
         across(BF, ~ifelse(.>100, "> 100", as.character(round(.,2))))) %>% 
  select(-lower, -upper) %>% 
  pivot_wider(names_from = param, values_from = c(mean, BF)) %>% 
  mutate(across(diffid, ~recode(., diff.1 = "before sentence vs word",
                                   diff.2 = "before vs within word"))) %>% 
  select(data, diffid, ends_with("beta"), ends_with("delta"), ends_with("theta"))

write_csv(tmp, "tables/location_effect_alldata.csv")

tmp %>% knitr::kable(align =c("l", "l", rep("r", 6)), 
                     col.names = c("Data set", "Difference", rep(c("Est. with 95% PIs", "BF"), 3)),
                     caption = "Effect of transition location on keystroke intervals. Differences between transition locations are shown on log scale (for transition durations) and logit scale for probability of hesitant transitions. 95% PIs in brackets.") %>%
  kableExtra::collapse_rows(columns = 1) %>%
  kable_styling(position = "center", full_width = T, font_size = 11) %>%
  footnote("PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.", threeparttable = T) %>% 
  add_header_above(c(" " = 2, 
                     "Fluent transitions" = 2, 
                     "Slowdown for hesitations" = 2, 
                     "Probability of hesitations" = 2))

@@

# Posterior by data set

## C2L1
